\section{Modelling First Order Differential Equations \& Autonomous Differential Equations}

\subsection{Modelling}

When modelling differential equations there are three important steps to follow:
\begin{enumerate}
    \item \bold{Translation}: this step involves understanding the problem, assigning variables to unknown quantities and developing equation(s) that describe the physical phenomena.
    \item \bold{Solving}: here, we decide whether to solve the system numerically or analytically, approximately or exactly, etc.. Our goal  here is to bring all information about the problem to bear.
    \item \bold{Interpret}: ensure that your answer you found in the previous step makes sense. Is your solution unique? Is it stable?
\end{enumerate}

\begin{example}[Radioactive Decay]
    Consider the following differential equation, which represents decay of a radioactive isotope:
    \[\frac{dN}{dt} = -\lambda N(t)\]
    In this example, $N(t)$ is the number of the radioactive isotope and $\lambda > 0$ is the decay constant. This equation can easily be solved using separation:
    \begin{align*}
        &\hspace{1cm}\frac{dN}{N(t)}= -\lambda dt \\
        & \implies \int \frac{dN}{N(t)} = \int -\lambda dt \\
        & \implies \ln|N(t)| = -\lambda t + C \\ 
        & \implies N(t) = Ae^{\lambda t}
    \end{align*}
    We remark that as $t \to \infty$, $N(t) \to 0$, which makes sense as the isotope is decaying.
\end{example}

\begin{example}[Newton's Law of Cooling]
    Suppose that we immerse a body in an environment with constant temperature $E$. Let $B(t)$ be the function representing the temperature of the body. We can represent the heat conductivity of the object with the equation:
    \[\frac{dB}{dt} = K(E - B(t))\]
    where $K > 0$ represents how well the body's material conducts heat. We can solve this equation by separation, which gives:
    \[B(t) = E - \frac{E - B_0}{e^{Kt}}\]
\end{example}

\begin{example}[Falling Bodies]
    We represent the fall of a body with mass $m$ with net forces $F(t, v)$ and acceleration $\frac{dv}{dt}$ using the differential equation:
    \[m \frac{dv}{dt} = F(t,v)\]
    For this class assume that downward displacements and forces are negative. Assume also that the force of gravity, $F_g = -mg$ where $g = 10 m/s^2$. 

    Additionally, we have kinematic equations for acceleratio $\mathbf{a}$, velocity $\mathbf{v}$, and displacement $\mathbf{r}$:
    \begin{align*}
        &\mathbf{a} = \frac{d\mathbf{v}}{dt} = \frac{d^2\mathbf{r}}{dt^2} \\
        &\mathbf{v} = \int \mathbf{a} dt = \mathbf{a}t + v_0 \\
        &\mathbf{r} = \int \mathbf{a}t + v_0 dt = \frac{1}{2}\mathbf{a}t^2 + v_0t + r_0
    \end{align*}
    We will use these equations to solve the following exercise:
\end{example}

\begin{example}
    A 50kg object is shot from a cannon straight up with an initial velocity of $10 m/s$ off the very tip of a bridge. If the air resistance is given by $5\mathbf{v}$ determine the velocity of the object at any time $t$.
\end{example}
\begin{solution}
    We have that:
    \[ m \frac{dv}{dt} = -10m - 5v \]
    Which can be written as:
    \[ \frac{dv}{dt} + \frac{5}{m}v = -10\]
    Let us solve this system using integrating factors. Let $\mu(t) = \exp \int p(t) \implies \mu(t) = \exp(\frac{5}{m}t)$. Multiplying the equation by the integrating factor gives:
\end{solution}
    \begin{align*}
        &\hspace{1.2cm} e^{5t/m}\frac{dv}{dt} + \frac{5}{m}e^{5t/m}v = -10e^{5t/m} \\
        &\iff \frac{d}{dt}\left[e^{5t/m}v\right] = -10e^{5t/m} \\
        &\iff e^{5t/m}v = \int -10e^{5t/m} dt \\
        &\iff v = -2m + Ce^{-5t/m}
    \end{align*}
    Since we know $v(0) = 0$ and $m = 50$, we have:
    \begin{align*}
        &\hspace{1cm} v(0) = 0 = -100 + C \\
        &\implies C = 100
    \end{align*}
    So the velocity of the object at time $t$ is:
    \[v(t) = -100 + 100e^{-t/10} \qedhere \]

\subsection{Autonomous Equations}

\begin{defbox}
    We say a first order DE is \bold{autonomous} if the independent variable does not appear explicitly in the equation. For example, the ODE $\frac{dy}{dt} = f(y)$ is autonomous as it depends on $y$ alone.
\end{defbox}

We have already seen some examples of autonomous ODEs in the first lecture, we will now classify certain two main types of autonomous ODEs:

\begin{defbox}
    We say an autonomous ODE is \bold{exponential} if it can be written in the form $\frac{dy}{dt} = ry$, where $r \neq 0$. If $r$ is positive, we say the equation represents \bold{exponential growth}, whereas if it is negative, the equation represents \bold{exponential decay}.
\end{defbox}

Now let us modify the exponential growth equation $\frac{dy}{dt} = ry$ such that $\frac{dy}{dt} \approx ry$ when $y$ is small, and $\frac{dy}{dt} \to 0$ as $y \to K$, where $K$ is the population that can be sustained. We write the equation as:
\[\frac{dy}{dt} = ry\left(1 - \frac{y}{K}\right)\]
where the term $\left(1 - \frac{y}{K}\right)$ is close to 1 when $y$ is small, and is close to 0 when $y \to K$. 

\begin{defbox}
    We say an autonomous ODE is \bold{logistic} if it can be written in the form $\frac{dy}{dt} = r\left(1 - \frac{y}{K}\right)y$, where $r \neq 0$, $K = r/a$ . If $r$ is positive, we say the equation represents \bold{logistic growth}, whereas if it is negative, the equation represents \bold{logistic decay}.
\end{defbox}

We say $r$ is the \textbf{intrinsic growth rate} of the system and $K$ is the system's \textbf{saturation level} or \textbf{carrying capacity}.

Observe that for small $y$, the rate of change is close to 1, but as $y$ tends to $k$, the rate of change approaches 0.

\begin{remark} 
   Solving the logistic equation (by using separation of variables) yields the following general solution:
   \[ y(t) = \frac{Ky_0}{y_0 + (K - y_0)e^{-rt}} \]
      Where $y_0$ is the value of $y(0)$. Observe that as $t \to \infty$, $y \to \frac{Ky_)}{y_0} = K$. 
\end{remark}
