\section{Linear \& Nonlinear ODEs, Integrating Factors}

\it{This lecture corresponds to \S1.3 and \S2.1 in the textbook.}

\subsection{Linear \& Nonlinear ODEs}

\begin{defbox}
    We say a differential equation $F(x, y, y', y'', \dots, y^{(n)}) = 0$ is \bold{linear} if $F$ is a linear function of $y, y', y'', \dots, y^{(n)}$. Thus the general form of a linear differential equation is:
    \[a_n(x)y^{(n)} + a_{n-1}(x)y^{(n-1)} + \dots + a_1(x)y' + a_0(x)y = g(x)\]
    We say an equation is \bold{nonlinear} if it is not linear.
\end{defbox}

The equation $y + y'' = 0$ is an example of a linear DE, as it can be written in the form $a_0(x)y + a_2(x)y'' = 0$ where $a_0(x) = a_2(x) = 1$. Another example of a linear DE is $y'\sin(x) = x$, where $a_1(x) = \sin(x)$ and $g(x) = x$. Conversely, the equation $y' + y^2 = 0$ is a nonlinear DE as it cannot be written in a linear form.

In mathematics, linear theory has been very well developed, but nonlinear theory has not been developed to the same degree. However, we have been able to approximate nonlinear equations by linear ones using \bold{linear approximations}.

For example, consider the following differential equation representing an oscillating pendulum:
\[\frac{d^2\theta}{dt^2} + \frac{g}{L}\sin\theta = 0\]
This equation is obviously nonlinear due to the $\sin\theta$. However, when the angle $\theta$ is small, we can approximate $\sin\theta \approx \theta$, hence we can linearize the equation as:
\[\frac{d^2\theta}{dt^2} + \frac{g}{L}\theta = 0\]
which is an equation we can solve (see exercises).

\subsection{Integrating Factors with Linear First Order ODEs}

In this section, we will be working with equations of the form $\frac{dy}{dx} + p(x)y = g(x)$, where $p$ and $g$ are functions of the independent variable $t$. We ,introduce a new method of solutions called integrating factors, with $\mu(t)$ being the integrating factor.

\begin{example}
    Solve the differential equation $y' + 2y = 1$ using integrating factors.
\end{example}
\begin{solution}
    The first step in solving an ODE with integrating factors is to multiply the entire equation by a function $\mu(t)$ which we will choose so the entire function is readily integrable. We have:
    \begin{align}
        \mu(t)\frac{dy}{dt} + \mu(t)2y = \mu(t)
    \end{align}
    The second step is to try and find the value of $\mu(t)$ such that this equation becomes readily integrable. We do this by considering the left side of Equation $(1)$ as the derivative of some recognizable expression. Observe that the term $\mu(t)\frac{dy}{dx}$ is part of the differentiating product $\mu(t)y$. Indeed if we recall the product rule: 
    \[\frac{d}{dt}\left[\mu(t)y\right] = \mu(t)\frac{dy}{dt} + \frac{d\mu}{dt}y\]
    we can find the integrating factor $\mu(t)$ by setting $\mu(t)2y = \frac{d\mu}{dt}y$. Isolating $\mu(t)$ we obtain:
    \begin{align*}
        &\hspace{1cm} \frac{1}{\mu(t)}d\mu = 2 dt \\
        &\implies \int \frac{1}{\mu(t)}d\mu = \int 2 dt \\
        &\implies \ln|\mu(t)| = 2t + C \\
        &\implies \mu(t) = e^{2t + C} \\
        &\implies \mu(t) = Ae^{2t}
    \end{align*}
    Hence Equation $(1)$ becomes (we can cancel out the constant $A$):
    \begin{align}
        e^{2t}\frac{dy}{dt} + e^{2t}2y = e^{2t}    
    \end{align}
    We remark that now the left side of Equation $(2)$ is obviously the derivative of the product $e^{2t}y$, so we can simplify the leftside furthermore to obtain:
    \[\frac{d}{dt}\left[e^{2t}y\right] = e^2t\]
    The third step is to integrate both sides of the Equation $(3)$, which we can now do with the chosen integrating factors, and isolate $y$. Remark that on the left side of the equation the derivative and integral cancel each other out:
    \begin{align*}
    &\hspace{1cm} \cancel{\int \frac{d}{dt}}\left[e^{2t}y\right] \cancel{dt} = \int e^{2t} dt \\
    &\implies e^{2t}y = \frac{e^2t}{2} + C \\
    &\implies y = \frac{1}{2} + Ce^{-2t} \qedhere
    \end{align*}
\end{solution}

\begin{remark}
    For ODEs of the form $\frac{dy}{dt} + p(t)y = g(t)$, the integrating factor is always of the form $\mu(t) = \exp \int p(t)dt$.
\end{remark}

\begin{example}
    Solve the differential equation $y' + 2ty = e^{-t^2}$ using integrating factors.
\end{example}
\begin{solution}
    We begin by multiplying the equation by the integrating factor $\mu(t)$:
    \begin{align}
        \mu(t)y' + \mu(t)2ty = \mu(t)e^{-t^2}
    \end{align}
    We first note that $p(t) = 2t$ and $g(t) = e^{-t^2}$. Using the remark above, we find that:
    \begin{align*}
        & \hspace{1.2cm} \mu(t) = \exp \int 2t dt \\ 
        &\iff \mu(t) = e^{t^2 + C} \\
        &\iff \mu(t) = Ae^{t^2}
    \end{align*}
    So Equation $(3)$ becomes:
    \[ e^{t^2}\frac{dy}{dt} + 2te^{t^2}y = 1 \]
    We recognize the left side of the equation as the derivative of the product $\mu(t)y$. Integrating both sides of the equation yields:
    \begin{align*}
        & \hspace{1cm} \cancel{\int \frac{d}{dt}}\left[e^{t^2}y\right] \cancel{dt} = \int dt \\
        & \implies e^{t^2}y = t + C \\
        & \implies y = e^{-t^2}(t + C) \qedhere
    \end{align*}
\end{solution}